Suppose the university is considering whether to run a late-night bus during exam period.

Let:

$$
\begin{aligned}
\theta
&= \text{the proportion of students who would use the bus}.
\end{aligned}
$$

### Prior

Before surveying students, we may not have a strong view about demand. It could be low, moderate, or high. For example, in the plot below:

- The $\text{Beta}(1,1)$ prior is flat. This means that before surveying students, we are treating low, moderate, and high levels of demand as equally plausible. 
- The centred prior, $\text{Beta}(5,5)$, places more weight near $\theta = 0.5$, which would represent a prior belief that demand is likely to be moderate. 
- The skewed prior, $\text{Beta}(5,2)$, places more weight on higher values of $\theta$, which would represent a prior belief that demand is more likely to be high than low.


```{r, warning=FALSE, echo=FALSE, message=FALSE}
library(tidyverse)

prior_df <- crossing(
  theta = seq(0.001, 0.999, length.out = 1000),
  prior = c("Flat: Beta(1, 1)", "Centred: Beta(5, 5)", "Skewed high: Beta(5, 2)")
) |>
  mutate(
    alpha = case_when(
      prior == "Flat: Beta(1, 1)" ~ 1,
      prior == "Centred: Beta(5, 5)" ~ 5,
      prior == "Skewed high: Beta(5, 2)" ~ 5
    ),
    beta = case_when(
      prior == "Flat: Beta(1, 1)" ~ 1,
      prior == "Centred: Beta(5, 5)" ~ 5,
      prior == "Skewed high: Beta(5, 2)" ~ 2
    ),
    density = dbeta(theta, shape1 = alpha, shape2 = beta),
    line_type = if_else(prior == "Flat: Beta(1, 1)", "Flat prior", "Other examples")
  )

ggplot(prior_df, aes(x = theta, y = density, group = prior)) +
  geom_line(
    data = prior_df |> filter(prior != "Flat: Beta(1, 1)"),
    aes(color = prior),
    linewidth = 1,
    alpha = 0.35
  ) +
  geom_line(
    data = prior_df |> filter(prior == "Flat: Beta(1, 1)"),
    aes(color = prior),
    linewidth = 1,
    alpha = 1
  ) +
  labs(
    x = expression(theta),
    y = "Prior density",
    color = NULL
  ) +
  scale_x_continuous(
    labels = scales::percent_format(),
    limits = c(0, 1)
  ) +
  scale_color_manual(
    values = c(
      "Flat: Beta(1, 1)" = "black",
      "Centred: Beta(5, 5)" = "blue",
      "Skewed high: Beta(5, 2)" = "darkgreen"
    )
  ) +
  theme_minimal()
```


To keep the example simple, we use a flat prior:

$$
\begin{aligned}
\theta
&\sim \text{Beta}(1,1)
\end{aligned}
$$

This prior treats all possible values of $\theta$ between 0 and 1 as equally plausible before seeing the survey data. 

### Sample

Now suppose we survey 10 students and 6 say they would use the bus:

$$
\begin{aligned}
x &= 6 \\
n &= 10
\end{aligned}
$$

This would produce a plot that looks like the following where the data peaks around $\theta=0.60$ because that is the observed sample proportion:

```{r}
likelihood_df <- tibble(
  theta = seq(0.001, 0.999, length.out = 1000),
  likelihood = dbinom(x = 6, size = 10, prob = theta),
  curve = "Likelihood: x = 6, n = 10"
)

ggplot(likelihood_df, aes(x = theta, y = likelihood, color = curve)) +
  geom_line(linewidth = 1) +
  labs(
    x = expression(theta),
    y = "Likelihood",
    color = NULL
  ) +
  scale_x_continuous(
    labels = scales::percent_format(),
    limits = c(0, 1)
  ) +
  scale_color_manual(
    values = c("Likelihood: x = 6, n = 10" = "red")
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```















